Formal verification of train route planning

### 2025-06-27 paper/main.tex
typos

\section{Introduction}
\label{sec:introduction}
\paragraph{Motivation.}

@@

\paragraph{Approach.}

To ensure the correctness of routes, we first formally model our system. Informally, our network is constituted of blocks, which can be used in both ways. Trains are given a list of orders, instructing them to go on one direction up to a certain block (modelling a form of odometer-based clearance), then changing direction, etc.. Beside these orders, trains respect signals they meet along the way. A regulator is in charge of setting traffic lights. We assume each block is equipped with reliable captors so that upon entering a block, an event is generated. The regulator react to those events asynchronously (to model lag). Regulator's reactions are used to update signals to release trains. Signals and captors are the only communications between trains and the regulator, and trains do not communicate directly with each other. Informally, our model is close to a multi-agent system or to communicating automata. 
After defining this model, we use TLA+~\cite{Lamport}, a specification tool and model checker suited for distributed systems, to model-check routes (i.e. combinations of train orders and regulator event handlers).

In the second part of this paper, we then write a compositional route generator, which aggregates individual conflict-free segments into general routes. We implement this compositional generator in a Python program, and we test it against our model checker.

@@

% Tools
As mentioned above, various works use formal methods for train systems. According to \cite{Ferrari_Formal_2022}, most works related to formal methods in railway systems use a combination of modelling and formal verification, e.g. model checking, used in approximately half of the works. The prime modelling language is B, used in 15\% of the works, well known for its industrial successes, e.g.\ for the development of the automated metro line 14 in Paris~\cite{butler_first_2020}. Ferrari et al.\ do not mention a single work using TLA+ for railway systems.

% Interlocking
One of the main challenges of both interlocking and route planning verification is composition: i.e. finding ways to cut a large problem into smaller parts, that are more easily solved, and which solutions are then assembled. \cite{haxthausen_automated_2023} propose a formally verified composition method for interlockings, using the RAISE Specification Language and a C++ implementation.

% Safety/Security
Finally, while most works focus on the safety of railway systems, we are seeing an increasing amount of works formally analysing the security of such systems in the context of cyberattacks. As a matter of fact, this contribution finds its origin in a cyber-warefare game organised for students at the University of Lorraine, see~\cite{CHE}. \cite{poorhadi_automating_2023} presents a interesting work modelling systems under cyber attacks in Event-B, helping designers to identify effects of attacks.

@@

Train events are handled by the regulator. Upon such events, the regulator can take actions such as turning a switch, allowing or blocking a train, etc. The regulator is equipped with a routing program. This routing program contains, for each train, a list of handlers. A handler is a tuple $\tuple{\bid{i}, A}$, where $\bid{i}$ is the position the train occupies when triggering the event, and $A$ is the list of orders associated with that event. For a given train, the first handler of the list is the set of actions that have to be executed on the train's first event, second handler for the second event, and so on.
Given a routing program $E$, we note $\handlerOf{E}{\tid{i}}$ the list of handlers of train $\tid{i}$, $\head{\handlerOf{E}{\tid{i}}} = \handlerOf{E}{\tid{i}}[1]$ the first handler of the list. Given a handler $h = \tuple{\bid{i}, a \cdot A}$, let  $\actionsOf{h} \triangleq a$, $\blockOf{h} \triangleq \bid{i}$ and $\nextAct{h} \triangleq \tuple{\bid{i}, A}$.  %(resp. \blockOf{h}) the list of action (resp. the block id) of $h$.
% We note $\nextAct{\handlerOf{E}{\tid{i}}}$ the routing program $E$ where the first event handler of train $\tid{i}$ is replaced with $\handler{\blockOf{\head{\handlerOf{E}{\tid{i}}}}}{\buftail{\actionsOf{\head{\handlerOf{E}{\tid{i}}}}}}$, i.e. where the first action of the first handler of train \tid{i} is removed.
 We note $\popHandlerHead{E}{\tid{i}}$ the routing program $E$ where the first event handler of train \tid{i} is removed, i.e. where $\handlerOf{E}{\tid{i}}$ is replaced by $\buftail{\handlerOf{E}{\tid{i}}}$.

@@

Then, the regulator lookup the first event on the buffer, and executes all the rules in the event handler associated with this event. All actions are treated in order, and when the list of remaining actions is empty, the handling terminates and the event is removed of the buffer. We introduce 6 reduction rules for the 4 possible kinds of orders (\incr{\bid{i}} and \wait{\bid{i}}{i} have two rules).

Rule \ruleRef{Turn} triggers when a \turn{\sid{i}}{s} is encountered in the event handler. It updates the state of the targeted switch.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Turn}]{
		\bufhead{\bufTrain} = \tid{i}
		\and
		\head{\handlerOf{E}{\tid{i}}} = \handler{\bid{i}}{\regConcat{\turn{\sid{i}}{s}}{A^\prime}}
	}{
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
	\\\reduces
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{i}}}}{T}{W}}{\updateSwitch{\switches}{\sid{i}}{s}}{\signals}}
	}
\end{mathpar}

@@

To execute a \wait{\bid{\text{crit}}}{n} order, the regulator stores the train id in its waiting set, together with the current block of the train (\bid{\text{src}}), the block it is waiting for (\bid{\text{crit}}), and the value the token associated with the target block has to reach to release the train ($n$).
\begin{mathpar}
	\inferrule*[left=\ruleDef{WaitBefore}] {
		\bufhead{\bufTrain} = \tid{w}
		\and
		\head{\handlerOf{E}{\tid{w}}} = \handler{\bid{\text{src}}}{\regConcat{\wait{\bid{\text{crit}}}{n}}{A^\prime}}
		\\
		\tokenOf{\bid{\text{crit}}} < n
	}{
		\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
		\\\reduces
		\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{w}}}}{T}{W\cup\tuple{\bid{\text{src}}, \bid{\text{crit}}, n, \tid{w}}}}{\switches}{\signals}}
	}
\end{mathpar}

Then, when the releasing train frees the critical block, three effects happen simultaneously: first, the token of the critical block is incremented; second the train waiting for the incremented token is released (we remove the correspond tuple from the waiting set)\footnote{By construction, nothing prevents multiple trains to be waiting for the same value, thus resulting in a faulty system. However, our model checker ensures programs (including event handlers) are correct, thus detecting such case.}; finally signals are updated to let the waiting train go until its next waiting point\footnote{We simply need to set a new red light at the next waiting point, the current red light will be turned green by signal rules, see \ruleRef{SignalSfty}.}. The predicate \nextWait{\cdot} looks-up the event handlers of a train (\tid{w} here) to find the next block (with id \bid{\text{stop}}) where the train is expected to wait.

\begin{mathpar}
	\inferrule*[left=\ruleDef{IncrAfter}] {
		\bufhead{\bufTrain} = \tid{r}
		\and
		\actionsOf{\head{\handlerOf{E}{\tid{r}}}} = \regConcat{\incr{\bid{\text{crit}}}}{A^\prime}
		\\
		\exists \bid{\text{src}}, \tid{w} \text{ s.t. } w = \tuple{\bid{\text{src}}, \bid{\text{crit}}, \tokenOf{\bid{\text{crit}}} + 1, \tid{w}} \in W
		\\
		\nextWait{\handlerOf{E}{\tid{w}}} = \bid{\text{stop}}
	}{
		\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
		\\\reduces
		{\begin{aligned}[t]
			\langle&\guardR, 
			\bufTrain,
			\push{\bufSig}{\tuple{\bid{\text{stop}}, \sigred}}, \\
			&\langle\trainSeq, \regTuple{\nextAct{\handlerOf{E}{\tid{r}}}}{\incrToken{T}{\bid{\text{crit}}}}{W\setminus\{w\}}, \switches, \signals\rangle\rangle
		\end{aligned}}
	}
\end{mathpar}

@@

Then, when the waiting train arrives, instead of stopping the train, the regulator can update the signal and let it continue until its next waiting point. This case is identified when the token already has the value the train expects\footnote{Notice that if the train misses its slot, e.g. if an incorrect program increases twice the token, the train will not continue, as another train, waiting for the twice-increased value, might be on the tracks.}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{WaitAfter}] {
		\bufhead{\bufTrain} = \tid{w}
		\and
		\actionsOf{\head{\handlerOf{E}{\tid{w}}}} = \regConcat{\wait{\bid{\text{crit}}}{n}}{A^\prime}
		\\
		\tokenOf{\bid{\text{crit}}} \geq n
		\and
		\nextWait{\handlerOf{E}{\tid{w}}} = \bid{\text{stop}}
	}{
		\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
		\\\reduces
		\redTuple{\guardR}{\bufTrain}{\push{\bufSig}{\tuple{\bid{\text{stop}}, \sigred}}}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{w}}}}{T}{W}}{\switches}{\signals}}
	}
\end{mathpar}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{Auth}] {
		\bufhead{\bufTrain} = \tid{i}
		\and
		\bufhead{E[\tid{i}]} = \handler{\bid{src}}{\regConcat{\auth}{A^\prime}}
		\\
		\nextWait{\buftail{E[\tid{i}]}} = \bid{\text{stop}}
	}{
			\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
			\\\reduces
			\redTuple{\guardR}{\bufTrain}{\push{\bufSig}{\tuple{\bid{\text{stop}}, \sigred}}}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{i}}}}{T}{W}}{\switches}{\signals}}
	}
\end{mathpar}
Notice that here, contrary to incr/wait rules, we consider the tail of the event handlers of the train, since it is already waiting and we want the next waiting location.

>>>>

\section{Introduction}
\label{sec:introduction}

This contribution finds its origin in a cyber-warefare we are organising for students at the University of Lorraine concerned by cyber-security, see~\cite{CHE}. In the game-play, we have both IT infrastructures and physical systems, with connections between them. On the physical side, we built a small train network that tries to be a good model of real systems.  Our system manages customer informations, train regulation and finally the physical/infrastructure layer. Attackers can hit any of the three parts: disfiguring the website, changing a platform or blocking a signal. Defenders will protect their system, typically via a strict network policy.

\paragraph{Routing trains.}
%\todo{Ajouter: tout est automatique}
Routes are computed on the fly. The schedule is renewed every "day"\footnote{We use a time scale factor: one of our "day" corresponds to 20 minutes in real time.}  and we introduce some randomness in the process: beyond some routine daily planning, some trains are inserted/deleted, some trains departure/destination may change, and so on. We simulate also delays. Finally, in our model, we distinguish freight and  passengers routes giving priority to the latter. Routes are recomputed each time a train has finished its duty. The general motivation of this paper is to propose a lightweight scheduling solution that is adapted to the fully automatic control, that takes into account the distributed nature of the system, that is reliable and does not suffer from interlocking.

We can now describe in more detail the working assumptions we are making about planning.  First, we have sensors on the rails at the two ends of each block. Accordingly,  we can catch train entrances within blocks. We make the hypothesis that this task is performed without errors. Second, the switches can be swapped from one position to the other, again, we suppose it is done without fault. Our method could cope such issues, but we leave that for future work.

  Third, for trains, we have a simple policy.  According to its plan, a train may start in some direction (Forward/Backward) up to some final position. Then, once started, the train will run according to the signals: when green, it (re-)starts, when red, it stops.  We do not make hypotheses on the durations of travels, but once a train is running, it will reach in finite time the next block. Trains run in parallel independently. One train may enter a block, while an other is crossing some switch and the third one is breaking. There is no communications between the trains.

Fourth,  the regulator is capable of reading train block entrance events, requiring some switch swap and finally locking (that is make sure a signal is red)/unlocking some signal. We consider that those latter operations can be carried out without delay. Finally, the regulator may append some new duty to a train. To sum up, to control the train mouvements, the regulator can only lock/unlock signals. Naturally, a signal intended to block train $t_1$ may block $t_2$ if not set at the right time.

% Parler de la stratégie, du modèle générale. All these operations are done by what we call  \emph{orders}. Roughly speaking, trains Orders have the shape "Start in direction Until position". Regulator orders are: %\todo{en parler}.

 To sum up, the regulation policy can be seen as an asynchronous distributed system. The N trains and the regulator are independent agents. That led us to TLA+, a framework introduced by Leslie Lamport for these kind of problems~\cite{Lamport}.  
%\todo{ Comparaison avec les travaux antérieurs / lit review}
Due to the distributed nature of the regulation, building a safe policy can be very tricky. We want to show in this paper that tools from formal methods are very useful in this respect. We describe In a first step, we describe a TLA+ specification that will allow us checking two properties of route orchestration: safety and liveness. This is our first research question.

Our second research question is the related to the following need. Suppose we want to append a new duty once a train reached its destination. This can be formalised as a composition problem (how to compose/concatenate some routes with some others). We describe such a composition procedure and then, again, we check that the routes are both safe and deadlock free.

For our last research problem, we show how one can build a global planning by means of composition of elementary steps. We check with TLA+ those compositions are correct.

\paragraph{Motivation.}

@@

@@

@@

Train events are handled by the regulator. Upon such events, the regulator can take actions such as turning a switch, allowing or blocking a train, etc. The regulator is equipped with a routing program. This routing program contains, for each train, a list of handlers. A handler is a tuple $\tuple{\bid{i}, A}$, where $\bid{i}$ is the position the train occupies when triggering the event, and $A$ is the list of orders associated with that event. For a given train, the first handler of the list is the set of actions that have to be executed on the train's first event, second handler for the second event, and so on.
Given a routing program $E$, we note $\handlerOf{E}{\tid{i}}$ the list of handlers of train $\tid{i}$, $\head{\handlerOf{E}{\tid{i}}} = \handlerOf{E}{\tid{i}}[1]$ the first handler of the list. Given a handler $h = \tuple{\bid{i}, A}$, let  $\actionsOf{h} \triangleq A$, $\blockOf{h} \triangleq \bid{i}$. When $h = \tuple{\bid{i}, a \cdot A}$ is the head of a sequence $H = h \cdot H'$,  $\nextAct{H} \triangleq \tuple{\bid{i}, A} \cdot H'$. %}{} \triangleq H'$ .  %(resp. \blockOf{h}) the list of action (resp. the block id) of $h$.
% We note $\nextAct{\handlerOf{E}{\tid{i}}}$ the routing program $E$ where the first event handler of train $\tid{i}$ is replaced with $\handler{\blockOf{\head{\handlerOf{E}{\tid{i}}}}}{\buftail{\actionsOf{\head{\handlerOf{E}{\tid{i}}}}}}$, i.e. where the first action of the first handler of train \tid{i} is removed.
% We note $\popHandlerHead{E}{\tid{i}}$ the routing program $E$ where the first event handler of train \tid{i} is removed, i.e. where $\handlerOf{E}{\tid{i}}$ is replaced by $\buftail{\handlerOf{E}{\tid{i}}}$.

@@

Then, the regulator lookup the first event on the buffer, and executes all the rules in the event handler associated with this event. All actions are treated in order, and when the list of remaining actions is empty, the handling terminates and the event is removed of the buffer. We introduce 6 reduction rules for the 4 possible kinds of orders (\incr{\bid{i}} and \wait{\bid{i}}{n} have two rules).

Rule \ruleRef{Turn} triggers when a \turn{\sid{i}}{s} is encountered in the event handler. It updates the state of the targeted switch.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Turn}]{
		\bufhead{\bufTrain} = \tid{i}
		\and
		\actionsOf{\head{\handlerOf{E}{\tid{i}}}} = \regConcat{\turn{\sid{i}}{s}}{A^\prime}
	}{
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
	\\\reduces
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{i}}}}{T}{W}}{\updateSwitch{\switches}{\sid{i}}{s}}{\signals}}
	}
\end{mathpar}

@@

To execute a \wait{\bid{\text{crit}}}{n} order, the regulator stores the train id in its waiting set, together with the current block of the train (\bid{\text{src}}), the block it is waiting for (\bid{\text{crit}}), and the value the token associated with the target block has to reach to release the train ($n$).
\begin{mathpar}
	\inferrule*[left=\ruleDef{WaitBefore}] {
		\bufhead{\bufTrain} = \tid{w}
		\and
		\head{\handlerOf{E}{\tid{w}}} = \handler{\bid{\text{src}}}{\regConcat{\wait{\bid{\text{crit}}}{n}}{A^\prime}}
		\\
		\tokenOf{\bid{\text{crit}}} \neq n
	}{
		\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
		\\\reduces
		\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{w}}}}{T}{W\cup\tuple{\bid{\text{src}}, \bid{\text{crit}}, n, \tid{w}}}}{\switches}{\signals}}
	}
\end{mathpar}

Then, when the releasing train frees the critical block, three effects happen simultaneously: first, the token of the critical block is incremented; second the train waiting for the incremented token is released (we remove the correspond tuple from the waiting set)\footnote{By construction, nothing prevents multiple trains to be waiting for the same value, thus resulting in a faulty system. However, our model checker ensures programs (including event handlers) are correct, thus detecting such case.}; finally signals are updated to let the waiting train go until its next waiting point. The predicate \nextWait{\cdot} looks-up the event handlers of a train (\tid{w} here) to find the next block (with id \bid{\text{stop}}) where the train is expected to wait.

\begin{mathpar}
	\inferrule*[left=\ruleDef{IncrAfter}] {
		\bufhead{\bufTrain} = \tid{r}
		\and
		\actionsOf{\head{\handlerOf{E}{\tid{r}}}} = \regConcat{\incr{\bid{\text{crit}}}}{A^\prime}
		\\
		\exists \bid{\text{src}}, \tid{w} \text{ s.t. } w = \tuple{\bid{\text{src}}, \bid{\text{crit}}, \tokenOf{\bid{\text{crit}}} + 1, \tid{w}} \in W
		\\
		\nextWait{\handlerOf{E}{\tid{w}}} = \bid{\text{stop}}
	}{
		\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
		\\\reduces
		{\begin{aligned}[t]
			\langle&\guardR, 
			\bufTrain,
			\push{\bufSig}{\tuple{\bid{\text{stop}}, \sigred}, \tuple{\bid{\text{src}}, \siggreen}}, \\
			&\langle\trainSeq, \regTuple{\nextAct{\handlerOf{E}{\tid{r}}}}{\incrToken{T}{\bid{\text{crit}}}}{W\setminus\{w\}}, \switches, \signals\rangle\rangle
		\end{aligned}}
	}
\end{mathpar}

@@

Then, when the waiting train arrives, instead of stopping the train, the regulator can update the signal and let it continue until its next waiting point. This case is identified when the token already has the value the train expects\footnote{Notice that if the train misses its slot, e.g. if an incorrect program increases twice the token, the train will not continue, as another train, waiting for the twice-increased value, might be on the tracks.}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{WaitAfter}] {
		\bufhead{\bufTrain} = \tid{w}
		\and
		\actionsOf{\head{\handlerOf{E}{\tid{w}}}} = \regConcat{\wait{\bid{\text{crit}}}{n}}{A^\prime}
		\\
		\tokenOf{\bid{\text{crit}}} = n
		\and
		\nextWait{\handlerOf{E}{\tid{w}}} = \bid{\text{stop}}
	}{
		\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
		\\\reduces
		\redTuple{\guardR}{\bufTrain}{\push{\bufSig}{\tuple{\bid{\text{stop}}, \sigred}, \tuple{\bid{\text{src}}, \siggreen}}}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{w}}}}{T}{W}}{\switches}{\signals}}
	}
\end{mathpar}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{Auth}] {
		\bufhead{\bufTrain} = \tid{i}
		\and
		\bufhead{E[\tid{i}]} = \handler{\bid{src}}{\regConcat{\auth}{A^\prime}}
		\\
		\nextWait{\buftail{E[\tid{i}]}} = \bid{\text{stop}}
	}{
			\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
			\\\reduces
			\redTuple{\guardR}{\bufTrain}{\push{\bufSig}{\tuple{\bid{\text{stop}}, \sigred}, \tuple{\bid{\text{src}}, \siggreen}}}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{i}}}}{T}{W}}{\switches}{\signals}}
	}
\end{mathpar}
Notice that here, contrary to incr/wait rules, we consider the tail of the event handlers of the train, since it is already waiting and we want the next waiting location.
